A Backsolving Technique Using Dynamic Taint Analysis

### 2013-12-13 docs/paper/fsuthesis/thesis-template/paper.tex
working on sigplan for netsec final report

>>>>

\section{Parallelization}

I have implemented a parallel version of the pintool, mp.c, to fork and exec a user-specified number of processes. Because a tainted branch may only be reachable based upon previous branches taken, and also because of complex conditions involving loops, the problem is not embarassingly parallel. For this project I have tested my program on a personally created binary, simpleCrackme. This binary takes a file containing a password in argv1. If the password matches its secret, then it prints "You win", otherwise, it prints "You lose".

After running my program, one can grep for the win string in the output/ directory. This is similar to concept to running a fuzzer, and grepping around the output recorded for more interesting conditions, like segfaults. Results show a noticeable speedup.

Sequential:
After several runs, I've seen performance ranging from ~33 seconds to 19 seconds.
26.12user 7.15system 0:33.42elapsed 99\%CPU (0avgtext+0avgdata 12964maxresident)k
0inputs+1496outputs (0major+1168365minor)pagefaults 0swaps

14.66user 4.67system 0:19.44elapsed 99\%CPU (0avgtext+0avgdata 12964maxresident)k
0inputs+1496outputs (0major+1168305minor)pagefaults 0swaps

Parallel:
The parallel implementation seems to decrease time by somewhere around half, which makes sense considering the problem is not embarassingly parellel, and Intel Pin may try and use multiple cores during startup.
30.64user 9.71system 0:10.83elapsed 372\%CPU (0avgtext+0avgdata 12952maxresident)k
0inputs+816outputs (0major+1368622minor)pagefaults 0swaps

However, I have seen runs as high as ~13 seconds elapsed. Since my testing environment has 4GB ram and 4 virtual cores, 4 processes seems to give me the best speedup. I also tested with 2, 8, and 16 processes but didn't see a comparable speedup.

### 2013-12-13 docs/paper/sigplanconf-template.tex
working on sigplan for netsec final report

Finding exploitable bugs in binaries is a difficult and lucrative endeavour. Fuzzing, or the feeding of controlled input into processes to try and detect vulnerable conditions [CITE], is one technique used by researchers and analysts. Applications tested for vulnerabilities are often black boxes because source code is unavailable. This piles difficulties onto an already complex problem, of which much of current research attempts to solve or reduce [CITES]. Among these issues are i) efficient and thorough dynamic taint analysis [CITE], ii) the path explosion problem [U], and iii) the code coverage problem [CITE?]. Dynamic taint analysis is difficult because taint can propagate directly via assignment, or indirectly via affecting control flows which affect assignment [D], which make proper implementation tricky. The path explosion problem arises from the nature of assembly languages. Jump instructions, which may make up as much as X\% of machine instructions in programs [CITE], each multiply the total number of existing paths in a binary by two. Considering the size of binaries, this quickly leads to an un-tenably large set of possible paths to explore via conventional techniques like symbolic execution. And even if it is possible to explore all paths in a binary, analysts still need sample input which can trigger the vulnerable condition, perhaps proving that the binary is exploitable along the way. Such automated exploit generation is the subject of much research [H, I, CITES].

This work aims to reduce the severity of the latter two problems while being wary of the first during implementation. Instead of trying to attain more reasonable performance by using symbolic execution less or more wisely [J], I focus instead on approximating symbolic execution by using dynamic taint analysis to drive mutation. This is possible by tracking taint at the instruction level, then backsolving from the jump to the source of taint to come up with smarter ways to mutate user input and create new test cases. For instance, in the figure 1 example:
\\
\\

@@

The major impetus for this work is outlined in "A Taint Based Approach for Smart Fuzzing". This paper proposes a fuzzing tool architecture, involving vulnerability detection, followed by taint analysis based upon found vulnerabilities, which drives the generation of intelligent tests. As tests are run they are checked to ensure they provide adequate code coverage, and results are carefully monitored for interesting situations like crashes [A]. The paper also serves as an excellent explanation of the current state and future of fuzzing. In particular, they suggest combining taint analysis with backward slicing. The technique discussed in the paper boils down to an attempt to automatically solve slices like those introduced in [AA], which have been derived from taint analysis as much as possible.

@@

I assume ASLR is disabled for the binary to be fuzzed. This allows the program to easily track and compare execution paths by creating a list of jump locations. Since the world is moving more and more toward 64-bit architectures, I plan to implement the technique for 32-bit x86 assembly. This is because x86 is widespread, and vulnerabilities resulting from unsafe C code should persist in binaries, regardless of both 32 vs. 64-bit and the flavor of architecture (ARM, x86, PowerPC), since compilers try to adhere to the C standard.

@@

I am hopeful about the results because current selective symbolic execution engines still appear to introduce significant overhead (Between 6 and 78X overhead more than QEMU for S2E. Just QEMU is between 4 and 10 times slower on some benchmarks, closer to 15 on others) [J, K, L]. In contrast, basic block counting using Intel Pin, the framework I am leveraging, introduced between 2 and 4X overhead [M]. However, the pintool needed for my proposed technique will introduce greater levels of overhead, since I will be working at the instruction level instead of the basic block level.

I am leveraging Intel PIN to perform dynamic binary instrumentation. I plan to test first on dummy programs with many paths that are easy to backsolve, similar to the example above. A C program which uses sequential switch statements to check user input against a secret, when compiled with tcc, which performs very little optimization, should create an appropriate binary. I then plan to test against S2E and Peach Fuzzer [Z]. I already have this program written and have verified using IDA Pro that there are a many, many paths to try and find.

There are allowable situations in programs which are outside the scope of this program. For instance, programs can loop forever [X]. These can be dealt with using reasonable heuristics and could be the subject of future research.

@@

Several valuable techniques to improve fuzzers' code coverage have evolved over the years. Two, in particular, inspire this work directly: Dynamic Taint Analysis and Concolic Execution. Dynamic Taint Analysis, or DTA, provides valuable information for guided fuzzing. Concolic execution, which has spawned out of the desire to use symbolic execution more wisely due to performance constraints, aims to ameliorate the path explosion problem. These each come with their own drawbacks, a primary one of which is the lack of automated instrumentation. Historically, symbolic execution required a programmer to directly instrument source code, but much research is now focused on automating as much as possible.

@@

Current research focuses on carefully deciding when and how to use symbolic execution [J, T, W, CITE]. So-called concolic execution mixes concrete and symbolic input, and various heuristics to determine when to resort to symbolic execution are in research [CITE].

Concolic testing has been shown to improve runtime while still allowing both wide and deep inspection of a program's execution tree, although it often still requires source code to be instrumented [X, Y]. What I propose is akin to concolic testing, in that I am augmenting concrete execution with dynamic taint analysis to solve for mutations which will exercise new paths in the code. However, the lack of a heavyweight constraint solver and the ability to work on binaries without instrumenting or using source code in any way have, to the best of my knowledge, not yet been proposed.

@@

Operations involving an untainted first operand and a tainted second operand generally spread taint, while operations with a tainted first operand and an untainted second operand remove taint. Tainted branches trigger the backsolving technique. Various precautions, such as removing taint for an instruction like "xor eax, eax", must be taken. Lossy operations, like shifts, must be handled carefully to be backsolve-able.

@@

Professional fuzzing products, like Peach Fuzzer

@@

Testing can begin on dummy programs which use a subset of all possible comparisons and arithmetic operations on tainted variables. If results appear promising I can try to make the fuzzer as complete as possible and begin running it on real-world applications. Achieving a noticeable speedup over modern symbolic execution would be the goal.

\subsection{Evaluation}

>>>>

Finding exploitable bugs in binaries is a difficult but valuable endeavour. Fuzzing, or the feeding of controlled input into processes to try and detect vulnerable conditions, is one technique used by researchers and analysts. Applications tested for vulnerabilities are often black boxes because source code is unavailable. This piles difficulties onto an already complex problem, of which much of current research attempts to solve or reduce. Among these issues are i) efficient and thorough dynamic taint analysis, ii) the path explosion problem [U], and iii) the code coverage problem. Dynamic taint analysis is also difficult because taint can propagate directly via assignment, or indirectly via affecting control flows which affect assignment [D], which make proper implementation tricky. The path explosion problem arises from the nature of assembly languages. Jump instructions, which may make up a decent percentage of machine instructions in programs, each multiply the total number of existing paths in a binary by two. Considering the size of binaries, this quickly leads to an un-tenably large set of possible paths to explore via conventional techniques like symbolic execution. And even if it is possible to explore all paths in a binary, analysts still need sample input which can trigger the vulnerable condition, perhaps proving that the binary is exploitable along the way. Such automated exploit generation is the subject of much research [H, I].

This work aims to reduce the severity of the latter two problems while being wary of the first during implementation. Instead of trying to attain more reasonable performance by using symbolic execution less or more wisely [J], I focus instead on approximating symbolic execution by using dynamic taint analysis to drive mutation. This is possible by tracking taint at the instruction level, then backsolving from the jump to the source of taint to determine smarter ways to mutate user input and create new test cases. For instance, in the figure 1 example:
\\
\\
\\

@@

The major impetus for this work is outlined in "A Taint Based Approach for Smart Fuzzing". This paper proposes a fuzzing tool architecture, involving vulnerability detection, followed by taint analysis based upon found vulnerabilities, which drives the generation of intelligent tests. As tests are run they are checked to ensure they provide adequate code coverage, and results are carefully monitored for interesting situations like crashes [A]. The paper also serves as an excellent explanation of the current state and future of fuzzing. In particular, they suggest combining taint analysis with backward slicing. The technique discussed in this paper boils down to an attempt to automatically solve slices like those introduced in [AA], which have been derived from taint analysis as much as possible. This data is put to use by solving for the input state necessary to explore new blocks in the binary by taking the opposite path for a particular branch in the previous run.

Often, research leveraging taint analysis requires access to source code or even programming optimization. However, minimizing programmer workload is more in spirit with fuzzing. Fuzzers should be fire and forget, allowing expensive programmers to work on something else while testing runs in the background. Work such as [E] has kept the spirit of this, although they focus at a different granularity of taint collection and propagation. The authors argue for following taint at the function level instead of at the instruction level, primarily to improve performance. The paper proposes several performance-saving heuristics, such as statically  constructing a finite state machine, which is executed with information gleaned during dynamic analysis, to help follow taint propagation within loops. However, perhaps because of the reliance on QEMU, or the initial translation of code to an intermediate representation without side-effects, significant overhead is introduced.

@@

I assume ASLR is disabled for the binary to be fuzzed. This allows the program to easily track and compare execution paths. I plan to implement the technique for 32-bit x86 assembly, because x86 is widespread, and vulnerabilities resulting from unsafe C code should persist in binaries, regardless of both the size and flavor of architecture (32 vs. 64-bit, ARM vs. x86 vs. PowerPC etc.), since compilers try to adhere to the C standard and vulnerabilities like buffer overflows exist at a C source code level.

@@

Because current selective symbolic execution engines still appear to introduce significant overhead (Between 6 and 78X overhead more than QEMU for S2E. Just QEMU is between 4 and 10 times slower on some benchmarks, closer to 15 on others) [J, K, L]. In contrast, basic block counting using Intel Pin, the framework I am leveraging, introduced between 2 and 4X overhead [M]. However, the pintool needed for my proposed technique will introduce greater levels of overhead, since I will be working at the instruction level instead of the basic block level, requiring more instrumentations for a given binary.

I am leveraging Intel PIN to perform dynamic binary instrumentation. I am first testing dummy programs with many paths that are easy to backsolve, similar to the example above. A C program which uses sequential switch statements to check user input against a secret, when compiled with tcc, which performs very little optimization, is such an example binary. I then plan to test against S2E and Peach Fuzzer [Z].

In addition, there are allowable situations in programs which are outside the scope of this program. For instance, programs can loop forever [X]. These can be dealt with using reasonable heuristics and could be the subject of future research.

@@

Several valuable techniques to improve fuzzers' code coverage have evolved over the years. Two, in particular, inspire this work directly: Dynamic Taint Analysis and Concolic Execution. Dynamic Taint Analysis, or DTA, provides valuable information for guided fuzzing. Concolic execution, which has spawned out of the desire to use symbolic execution more wisely due to performance constraints, aims to ameliorate the path explosion problem while maintaining good code coverage. These each come with their own drawbacks, a primary one of which is the lack of automated instrumentation. Historically, symbolic execution required a programmer to directly instrument source code, but much research is now focused on automating as much as possible.

@@

Current research focuses on carefully deciding when and how to use symbolic execution [J, T, W]. So-called concolic execution mixes concrete and symbolic input, and various heuristics to determine when to resort to symbolic execution are in research.

Concolic testing has been shown to improve runtime while still allowing both wide and deep inspection of a program's execution tree, although it often still requires source code to be instrumented [X, Y]. What I propose is similar to concolic testing, in that I am augmenting concrete execution with dynamic taint analysis to solve for mutations which will exercise new paths in the code. However, the lack of a heavyweight constraint solver and the ability to work on binaries without instrumenting or using source code in any way have, to the best of my knowledge, not yet been proposed.

@@

For example, move operations involving an untainted first operand and a tainted second operand generally spread taint, while those with a tainted first operand and an untainted second operand remove taint. Tainted branches trigger the backsolving technique. Various precautions, such as removing taint for an instruction like "xor eax, eax", must be taken. Lossy operations, like shifts, must be handled carefully to be backsolve-able.

@@

%Professional fuzzing products, like Peach Fuzzer

@@

I am currently instrumenting dummy programs which use a subset of all possible comparisons and arithmetic operations on tainted variables. If results appear promising I can try to make the fuzzer as complete as possible and begin running it on real-world applications. Achieving a noticeable speedup over modern symbolic execution would be the goal.

As of this writing I am moving towards fuzzing Linux utilities. Results for toy binaries are promising, with "crackme" type programs which ask for a secret being solved in approximately 2*n runs, where n is the length of the secret. There is, however, still some overtainting which must be accounted for.	
%\section{Evaluation}

\subsection{Parallelization}

I have implemented a parallel version of the pintool, mp.c, to fork and exec a user-specified number of processes. Because a tainted branch may only be reachable based upon previous branches taken, and also because of complex conditions involving loops, the problem is not embarassingly parallel. For this project I have tested my program on a personally created binary, simpleCrackme. This binary takes a file containing a password in argv1. If the password matches its secret, then it prints "You win", otherwise, it prints "You lose".

After running my program, one can grep for the win string in the output/ directory. This is similar to concept to running a fuzzer, and grepping around the output recorded for more interesting conditions, like segfaults. Results show a noticeable speedup.

Sequential:
After several runs, I've seen performance ranging from ~33 seconds to 19 seconds.
26.12user 7.15system 0:33.42elapsed 99\%CPU (0avgtext+0avgdata 12964maxresident)k
0inputs+1496outputs (0major+1168365minor)pagefaults 0swaps

14.66user 4.67system 0:19.44elapsed 99\%CPU (0avgtext+0avgdata 12964maxresident)k
0inputs+1496outputs (0major+1168305minor)pagefaults 0swaps

Parallel:
The parallel implementation seems to decrease time by somewhere around half, which makes sense considering the problem is not embarassingly parellel, and Intel Pin may try and use multiple cores during startup.
30.64user 9.71system 0:10.83elapsed 372\%CPU (0avgtext+0avgdata 12952maxresident)k
0inputs+816outputs (0major+1368622minor)pagefaults 0swaps

However, I have seen runs as high as ~13 seconds elapsed. Since my testing environment has 4GB ram and 4 virtual cores, 4 processes seems to give me the best speedup. I also tested with 2, 8, and 16 processes but didn't see a comparable speedup.
